Supporting Information

### 2016-08-15 manuscript.tex
ms edits, figure S3

To consider the joint effects of biased sex ratios and variance in male reproductive success, our neutrally-expected levels of diversity were calculated using equations 1-3, assuming no variance in female mating (e.g., $C^2_{f}=0$), and we considered a range of values for $\sigma_{m}$: 0 (no variance), 1/2 (Poisson/2) 1 (Poisson distribution of offspring), 2 (Poisson*2), and 3 (Poisson*3). Assuming that the level of neutral polymorphism is given by the product of the mutation rate and the effective population size, $\theta=4N_{e}\mu$ \citep{watterson1975,kimura1984}, Figure 1 shows the predicted effective population size of autosomes and sex chromosomes as a function of the sex ratio and variance in male reproductive success. Using these neutral predictions, we compared our empirically estimated diversity levels to normalized X/A and Y/A ratios of diversity. Note that we have only considered variance in male reproductive success here because this is most likely to occur in wind pollinated plant like \textit{R. hastatulus}, where males produce copious amounts of pollen and compete to fertilize uniovulate females (\X ref).

@@

\subsection*{Simulations of positive and purifying selection}
To test whether our observed level of Y-chromosome diversity could be explained by selection, we conducted two sets of forward-time simulations using the software SFSCODE \citep{hernandez2008flexible}: (i) Purifying selection acting to remove deleterious mutations and linked neutral variation (i.e., background selection), and (ii) Positive selection acting on beneficial mutations and reducing variation through selective sweeps (refs \X). For purifying selection simulations, we assumed that selection coefficients followed a gamma distribution, and we considered a range of values for the mean selection coefficient ($s$) by varying the scale parameter, $k$ (range \X) from (\X) to \X and keeping the shape parameter, \X, constant (see Supporting Information for simulation code). To make the simulation output comparable to our data, we initialized simulations with our empirically estimated value of $\theta$ (estimated from autosomes), and sampled 8 chromosomes per simulation from 24 populations. For simulating positive selection, we used \X [parameters here]. We ran 1000 replicates for each model and calculated the average $\theta_{sim}$ and the corresponding Y/A ratio of neutral diversity under the both sets of models separately (purifying and positive selection). We then tested the fit of our simulated diversity output to our empirical data (\X new figure) by calculating the proportion of simulation replicates for which $\theta_{sim}$ was not significantly different from $\theta_{observed}$, using a two-sided P value. Commands for running the simulations are provided in the Supporting Information.

@@

\subsection*{Y-chromosome diversity in \textit{R. hastatulus} is very low}
Our analysis revealed that diversity on the \textit{R. hastatulus} Y chromosome is significantly lower than expected under neutrality, with estimates indicating $Y/A=0.02$, which is 12.5 fold lower than the standard neutral prediction of $Y/A = 0.25 (P<0.0001)$, and a 40-fold reduction compared to mean diversity on the X chromosome (Table 1). Note that by normalizing X and Y diversity by autosomal diversity, our results indicate that the X-Y difference we observed was not due to an elevation of X chromosome diversity, but rather a Y-specific reduction. Conceivably, such low diversity on the Y could arise from a low mutation rate on the Y chromosome, or a lower mutation rate in males compared to females. However, these possibilities can be rejected because there is no evidence that the number of synonymous mutations in X and Y lineages, estimated by both parsimony and maximum likelihood, are significantly different \citep{hough2014}. It is worth noting that our results only apply to the $Y_{1}$ chromosome, however, as estimates of diversity were calculated using the set of sex-linked genes that were shared between the $XX/XY$ and the $XX/XY_{1}Y_{2}$ sex chromosome races. Previous work investigating patterns of genetic degeneration in both systems found that both the $Y_{1}$ and the more recently evolved "neo" $Y_{2}$ chromosomes have undergone classic signs of degeneration, including gene loss, loss of expression, and an accumulation of amino acid-changing mutations, but we cannot say at present whether the neo-Y has also undergone a significant reduction in diversity. Nonetheless, our results indicate that the pattern of reduced diversity on the $Y_{1}$ chromosome occurs in each of the \textit{R. hastatulus} sub-clades that were inferred to be monophyletic (\X Figure 2; Table 1; Figure S1,S2), suggesting that this effect is not population-specific.

@@

Although our sampling of each \textit{R. hastatulus} sub-clade is limited, the discovery of three phylogenetically distinct monophyletic groups is interesting as it suggests the possibility that introgression occurred between the ancestral Texas (XY race) and the derived North Carolina ($XY_{1}Y_{2}$) race, leading to a derived $XY_{1}Y_{2}$ sub-clade. As the North Carolina and Texas races are known to be inter-fertile \citep{smith1964evolving}, we suggest that the SC sub-clade inferred here likely originated through hybridization between a female from the FL clade harboring the X-autosome fusion, and a male from the XY Texas race. Notably, when we pooled the samples from each sub-clade together, the estimated level of Y diversity was significantly higher in the pooled data, highlighting the presence of strong substructure between these groups (Figure S2).

%Our data indicate a significant reduction in X/A polymorphism in the derived SC and FL sub-clades of the North Carolina race (X/AFL=0.33 and X/ASC=0.37) compared to the Texas race (X/ATX=0.85). This reduction in diversity may be associated with the recent origin of the XY1Y2 sex chromosome system that occurs in these two sub-clades. The XY1Y2 sex chromosome system is thought to have originated through an X-autosome fusion (Smith 1964) involving the ancestral 3rd chromosome in the Texas race. Evidence supporting this autosomal origin was recently obtained by Grabowska-Joachimiak et al. (2014), who reported that the ancestral third chromosome in the Texas race carries the 5S rDNA locus, which is now found on both the neo-X and the Y2 sex chromosomes in the derived North Carolina race. That this fusion is likely to have been recent is suggested by C-banding/DAPI experiments showing that while the ancestral Y chromosome is highly heterochromatic, the translocated neo-Y is still euchromatic (Grabowska-Joachimiak et al. 2014), and also shows less genetic degeneration than the ancestral Y chromosome (Hough et al. 2014). Therefore, if recent positive selection was responsible for driving the evolution of this fusion, then the formerly autosomal segment on the X chromosome in the XY1Y2 sub-clades (the neo-X) may have experienced a selective sweep. This process may explain the reduction in diversity that we observed in the derived XY1Y2 sub-clades, though further work is needed to determine the factors driving the establishment of the X-autosome fusion.

%We next consider several possible models that could explain this unexpectedly low amount of diversity found on the Y chromosome relative to other genomic regions. Such models include differences in the variance in reproductive success between males and females, purifying selection on the Y chromosome, and positive selection on the Y chromosome.

@@

 A highly reduced level of Y chromosome variability is a predicted outcome of models Y-chromosome evolution based on the removal of deleterious mutations and linked neutral variants (e.g., background selection), ...and several cases.... but the occurrence of female-biased sex ratios in this species has also been predicted to lower Y diversity. Moreover, a further reduction in Y diversity could occure because of high variance in male reproductive success (Figure 1). This is not unusual in annual species such as \textit{R. hastatulus}s, which commonly exhibit extensive phenotypic plasticity in plant size and flower production (\X Harper 1977). Conceivably, such plastic differences among plants, owing to spatial heterogeneity in resources, could influence the intensity of male-male competition, and thus affect variance in male reproductive success. Given that male plants in this wind-pollinated species produce large amounts of pollen, and that female flowers are uniovulate, one might expect strong competition among males.

@@

 %e.g., X/A diversity ratio that is 30% larger than what we found for the Texas race (X/ATX=0.85), and 80% larger than  the X/A ratio in the North Carolina race (X/ANC=0.35, further discussed below).

%In contrast to the Texas sex chromosome race, our analysis revealed that Y-chromosome diversity was considerably higher in the North Carolina race (see Supplementary Material; Figure S1). This difference could have arisen due to population substructure, resulting in an elevated overall Y diversity in the pooled sample. To test this hypothesis, we conducted a phylogenetic analysis of sex-linked sequences from all populations in our study. The resulting sequence phylogeny (Figure 4) revealed strong support for the Y-linked genes from the XY1Y2 race being paraphyletic (98% bootstrap support). In particular, our analysis revealed that samples from two populations, both from South Carolina (hereafter the SC sub-clade), formed a monophyletic group, whereas samples from Florida and Georgia (hereafter the FL sub-clade) were also monophyletic, but were more closely related to the XY Texas race. Although our sampling here is limited, these results suggest that introgression may have occurred between the ancestral Texas (XY race) and the derived North Carolina (XY1Y2) race, leading to a second, derived, XY1Y2 sub-clade and an elevated level of Y-diversity in our pooled sample. As these races are inter-fertile (Smith reference about hybrid zone), we suggest that the SC sub-clade likely originated through hybridization between a female from the FL clade harboring the X-autosome fusion, and a male from the XY race. Notably, when we separated samples by the two inferred sub-clades of Y–linked sequences, the estimated level of Y diversity was consistent across all three clades, suggesting that the extensive reduction in diversity on this chromosome is not a population or race-specific phenomenon (see Supplementary Material, Figure S1).

@@

\section*{Conclusions}
\section*{Acknowledgments}

>>>>

To consider the joint effects of biased sex ratios and variance in male reproductive success, our neutrally-expected levels of diversity were calculated using equations 1-3, assuming no variance in female mating (e.g., $C^2_{f}=0$), and we considered a range of values for $\sigma_{m}$: 0 (no variance), 1/2 (Poisson/2) 1 (Poisson distribution of offspring), 2 (Poisson*2), and 3 (Poisson*3). Figure 1 shows the predicted effective population size of autosomes and sex chromosomes as a function of the sex ratio and variance in male reproductive success. Using these neutral predictions, we compared our empirically estimated diversity levels to normalized X/A and Y/A ratios of diversity, assuming that the level of neutral polymorphism is given by the product of the mutation rate and the effective population size, $\theta=4N_{e}\mu$ \citep{watterson1975,kimura1984}. Note that we have only considered variance in male reproductive success here because this is most likely to occur in wind pollinated plant like \textit{R. hastatulus}, where males produce copious amounts of pollen and compete to fertilize uniovulate females (\X ref).

@@

\subsection*{Simulations of positive and purifying selection}
To test whether our observed level of Y-chromosome diversity could be explained by selection, we conducted two sets of forward-time simulations using the software SFSCODE \citep{hernandez2008flexible}: (i) Purifying selection acting to remove deleterious mutations and linked neutral variation (i.e., background selection), and (ii) Positive selection acting on beneficial mutations and reducing variation through selective sweeps (refs \X). For purifying selection simulations, we assumed that selection coefficients followed a gamma distribution, and we considered a range of values for the mean selection coefficient ($s$) by varying the scale parameter, $k$ (range \X) from (\X) to \X and keeping the shape parameter, \X, constant (see Supporting Information for simulation code). To make the simulation output comparable to our data, we initialized simulations with our empirically estimated value of $\theta$ (estimated from autosomes), and sampled 8 chromosomes per simulation from 24 populations. For simulating positive selection, we used \X [parameters here]. We ran 1000 replicates for each model and calculated the average $\theta_{sim}$ and the corresponding Y/A ratio of neutral diversity under the both sets of models separately. We then tested the significant of the fit of our simulated diversity output to our empirical data (\X new figure) by calculating the proportion of simulation replicates for which $\theta_{sim}$ was not significantly different from $\theta_{observed}$, using a two-sided P value. Commands for running the simulations are provided in the Supporting Information.

@@

\subsection*{Y-chromosome diversity in \textit{R. hastatulus} is very low}
Our analysis revealed that diversity on the \textit{R. hastatulus} Y chromosome is significantly lower than expected under neutrality, with estimates indicating Y/A=0.02, which is 12.5 fold lower than the standard neutral prediction of Y/A = 0.25 (P<0.0001), and 40-fold lower compared to mean diversity on the X chromosome (Table 1). Note that by normalizing X and Y diversity by autosomal diversity, our results indicate that the X-Y difference we observed was not due to an elevation of X chromosome diversity, but rather a Y-specific reduction. Conceivably, such low diversity on the Y could arise from a low mutation rate on the Y chromosome, or a lower mutation rate in males compared to females. However, these possibilities can be rejected because there is no evidence that the number of synonymous mutations in X and Y lineages, estimated by both parsimony and maximum likelihood, are significantly different \citep{hough2014}.

It is worth noting that our results only apply to the $Y_{1}$ chromosome, however, as our estimates of diversity were calculated using the set of sex-linked genes that were shared between the $XX/XY$ and the $XX/XY_{1}Y_{2}$ sex chromosome systems. Previous work investigating patterns of genetic degeneration found that both the $Y_{1}$ and the more recently evolved "neo" $Y_{2}$ chromosomes exhibit signs of genetic degeneration, including gene loss, loss of expression, and an accumulation of amino acid-changing mutations \citep{hough2014}, but we cannot say from the present study whether the neo-Y chromosome has also undergone a significant reduction in diversity. However, our results do indicate that the pattern of reduced diversity on the $Y_{1}$ chromosome occurs in each of the three \textit{R. hastatulus} sub-clades (\X Figure 2; Table 1; Figure S2), suggesting that this effect is not population-specific.

@@

Although our sampling of each \textit{R. hastatulus} sub-clade is limited, the discovery of three phylogenetically distinct monophyletic groups is interesting because it suggests the possibility that introgression occurred between the ancestral Texas (XY race) and the derived North Carolina ($XY_{1}Y_{2}$) race, leading to a derived $XY_{1}Y_{2}$ sub-clade. As the North Carolina and Texas races are known to be inter-fertile \citep{smith1964evolving}, we suggest that the SC sub-clade inferred here likely originated through hybridization between a female from the FL clade harboring the X-autosome fusion, and a male from the XY Texas race. Notably, when we pooled the samples from each sub-clade together, the estimated level of Y diversity was significantly higher in the pooled data, highlighting the presence of strong substructure between these groups (Figure S2).

Interestingly, our results indicate a significant reduction in X/A diversity in the derived SC and FL sub-clades of the North Carolina race ($X/A_{FL}=0.33$ and $X/A_{SC}=0.37$) compared to the Texas race ($X/A_{TX}=0.85$) (Figure 2). Although not expected, this reduction in diversity may be associated with the recent origin of the XY1Y2 sex chromosome system. The XY1Y2 sex chromosome system is thought to have originated through an X-autosome fusion (Smith 1964) involving the ancestral 3rd chromosome in the Texas race. Evidence supporting this autosomal origin was recently obtained by Grabowska-Joachimiak et al. (2014), who reported that the ancestral third chromosome in the Texas race carries the 5S rDNA locus, which is now found on both the neo-X and the Y2 sex chromosomes in the derived North Carolina race. If recent positive selection was responsible for driving the evolution of this fusion, then the formerly autosomal segment on the X chromosome in the XY1Y2 sub-clades (the neo-X) may have experienced a selective sweep, resulting in reduced X-linked diversity in the derived XY1Y2 sub-clades. It will be important for future work to investigate the factors driving the establishment of the X-autosome fusion in this species, and how they might impact patterns of X-linked neutral diversity.

In contrast to the situation on the X chromosome, our data indicate a strong and consistent diversity reduction on the Y chromosome, and we now test several possible models that might explain this diversity reduction.

@@

 A highly reduced level of Y chromosome variability is a predicted outcome of models Y-chromosome evolution based on the removal of deleterious mutations and linked neutral variants (e.g., background selection) (\X refs), ...and several cases....(\X refs), but the occurrence of female-biased sex ratios in this species has also been predicted to lower Y diversity in the absence of such selective interference. Moreover, such a reduction in Y-linked diversity could be accented due to high variance in male reproductive success (Figure 1). This is not unusual in annual plants such as \textit{R. hastatulus}, which commonly exhibit extensive phenotypic plasticity in plant size and flower production (\X Harper 1977). Conceivably, such plastic differences among plants, owing to spatial heterogeneity in resources, could influence the intensity of male-male competition, and thus affect variance in male reproductive success. Given that male plants in this wind-pollinated species produce large amounts of pollen, and that female flowers are uniovulate, strong competition among males is expected.

@@

@@

\section*{Conclusions}

\section*{Acknowledgments}

### 2016-08-15 supporting-information/supporting_info.tex
supp info edits, ms edits
ms edits, figure S3

\templatetype{pnassupportinginfo}

\title{Supporting Information}
\author{Hough et al.}

@@

\section*{SI Datasets}

>>>>

\templatetype{pnassupportinginfo}
\usepackage{listings}
\lstset{basicstyle=\ttfamily,
  showstringspaces=false,
  commentstyle=\color{red},
  keywordstyle=\color{blue}
}
\title{Supporting Information}
\author{Hough et al.}

@@

\begin{figure}[tbhp!]
\centering
\includegraphics[width=1\linewidth]{FigureS3.png}
\caption{Normalized X/A and Y/A nucleotide diversity ratios for XY and $XY_{1}Y_{2}$ races of \textit{Rumex hastatulus}. In \textbf{A}, the South Carolina and Florida sub-clades are collapsed into a single race (‘NC’), and in \textbf{B}, they are separated into the inferred ‘SC’ and ‘FL’ sub-clades.}
\label{figure:FigureS3}
\end{figure}

%Figure A4.1.2: Autosomal nucleotide diversity for the TX, SC, and FL clades of Rumex hastatulus, with confidence intervals calculated from bootstrapping 20000 replicates using the BCa method (Efron 1987) implemented in the Boot package in R (Canty and Ripley 2012; R Core Team 2011).

\newpage
\section*{Simulation code}

\begin{lstlisting}
#!/bin/bash

#purifying selection
sfs_code 1 50 -r 0 -t 0.001375 -P 1 -TE 0.3 -L 2 500000 800000 \
-a N -W L 1 2 0 1 1 0.184 0.003 -n 8 -N 500 -A

#positive selection
sfs_code 1 50 -r 0 -t 0.001375 -P 1 -TE 0.3 -L 2 500000 800000 \
-a N -W L 1 2 0 1 1 0.184 0.003 -n 8 -N 500 -A
\end{lstlisting}
